Fractional Types

### 2013-02-20 frac.tex
cleaned up a few small things

The extension of the sets of types and values is simple:

@@

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}. There is
however have some informal intuition about the \emph{meaning} of such
fractional types and values. (See Sec.~\ref{sec:cat} for more discussion
about this point.) From both the programming perspective and from the
categorical perspective we expect the fractional types to satisfy the
following isomorphism witnessed by two new combinators {{eta*_b}} and
{{eps*_b}}:

@@

From a programming perspective, we think of a value {{1/v}} as a
\emph{first-class constraint} that can only be satisfied if it is matched
with an actual value {{v}} or as a \emph{pattern} representing the absence of
some information of a specific shapre (i.e., \emph{negative information}),
that can only be reconciliated by an actual value {{v}}. The combinator
{{eta*_b : 1 <-> 1/b * b}} thus represents a fission point which creates ---
out of no information --- an equal amount of negative and positive
information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}} matches
up an equal amount of negative and positive information producing no residual
information. From the categorical perspective, the extended language {{PiTF}}
needs to have a semantics in a \emph{compact closed category} which are the
categories introduced by Abramsky and
Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
computation. Two of the most relevant features of these categories for our
purposes are that they support an adjoint for each morphism (to model
reversibility) and that they support higher-order relation (built from the
categorical dagger which is modeled by our fractional types). The main
requirement on the dagger operation is the isomorphism above witnessed by
{{eta*_b}} and {{eps*_b}}.

One can think of several concrete compact closed categories: the first and
most intuitive is the category of finite sets and relations. In that case,
the dagger operation (i.e., the fractional type) collapses by defining
{{1/b}} to be {{b}}. This is still an interesting category as any relation is
trivially reversible and more importantly any relation between {{b1}} and
{{b2}} can be represented as a subset of {{b1 * b2}}, i.e., as a value. A
more interesting example is the category of finite dimensional vector spaces
and linear maps (or even the category of finite dimensional Hilbert
spaces). In such categories the dagger operation takes a vector space
(consisting of ``column vectors'') to the dual vector space (consisting of
``row vectors''). These vector spaces are isomorphic, but they are
nevertheless quite different, being $(1,0)$ and $(0,1)$ tensors
(respectively).  In other words, the category provides a more refined model
in which isomorphic negative-information values and positive-information are
not identified.

In this section, we provide a simple semantics of {{PiTF}} in the category of
finite sets and relations. In other words, we interpret each combinator 
{{c :  b1 <-> b2}} as a \emph{relation} between the sets denoted
by {{b1}} and {{b2}} respectively. As we illustrate, this semantics allows
\emph{arbitrary} relations (including the empty relation and
relations that are not isomorphisms) to be represented as values in {{PiTF}}.
In Sec.~\ref{sec:cat} we discuss a more refined semantics that is
more suitable for richer categories and which only allows relations
representing \emph{isomorphisms} to be represented as values in~{{PiTF}}.

@@

The operational semantics define a relation rather than a function, now, but the
goal is for them to still be reversible; in particular, we wish for the relation
to mode check so that for any combinator {{c}} we can treat either its left or
its right side as an input; this allows us to define {{sym}} as just an argument
swap. In the interest of brevity, we have treated the type annotations in the
base isomorphisms as implicit; for example, the rule for {{assocr*}} should
really read

@@

The relational semantics can be given an operational interpretation by
defining two functions: a forward evaluator and a backwards evaluator that
exercise the relation in opposite directions. It follows that if the forward
evaluation of {{c v1}} produces {{v2}} as a possible answer then the
backwards evaluation of {{c v2}} produces {{v1}} as a possible answer.

@@

>>>>

The extension of the sets of types and values from {{Pi}} to {{PiTF}} is
simple:

@@

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}. We have
however some informal intuition about the \emph{meaning} of such fractional
types and values. (See Sec.~\ref{sec:cat} for more discussion about this
point.) In more detail, given a symmetric monoidal category modeling a
certain language (like {{Pi}}), one can obtain a higher-order extension of
the language by extending the category to a \emph{compact closed} one, i.e.,
a category in which morphisms are representable as objects. The key to this
extension is to add dual objects (in our case the fractionals) satisfying the
following isomorphism witnessed by two new combinators {{eta*_b}} and
{{eps*_b}}:

@@

\noindent From a programming perspective, we think of a value {{1/v}} as a
\emph{first-class constraint} that can only be satisfied if it is matched
with an actual value {{v}} or as a \emph{pattern} representing the absence of
some information of a specific shape (i.e., \emph{negative information}),
that can only be reconciliated by an actual value {{v}}. The combinator
{{eta*_b : 1 <-> 1/b * b}} thus represents a fission point which creates ---
out of no information --- an equal amount of negative and positive
information. Symmetrically, the combinator {{eps*_b : 1/b * b <-> 1}} matches
up an equal amount of negative and positive information producing no residual
information.

From the categorical perspective, versions of compact closed categories were
introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} as the
generalization of monoidal categories to model quantum computation.  The
first and most intuitive example of such categories is the category of finite
sets and relations. In that case, the dual operation (i.e., the fractional
type) collapses by defining {{1/b}} to be {{b}}. In spite of this degenerate
interpretation of fractionals, this is still an interesting category as it
provides a model for a reversible higher-order language. Indeed, any relation
is trivially reversible and any relation between {{b1}} and {{b2}} can be
represented as a subset of {{b1 * b2}}, i.e., as a value. A more interesting
example is the category of finite dimensional vector spaces and linear maps
(or even the category of finite dimensional Hilbert spaces). In such
categories the dual operation takes a vector space (consisting of ``column
vectors'') to the dual vector space (consisting of ``row vectors''). These
vector spaces are isomorphic, but they are nevertheless quite different,
being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words, the
category provides a more refined model in which isomorphic
negative-information values and positive-information are not identified.

In this section, we provide a simple semantics of {{PiTF}} in the category of
finite sets and relations. In other words, we interpret each combinator {{c :
    b1 <-> b2}} as a \emph{relation} between the sets denoted by {{b1}} and
{{b2}} respectively. As we illustrate, this semantics allows \emph{arbitrary}
relations (including the empty relation and relations that are not
isomorphisms) to be represented as values in {{PiTF}}.  In Sec.~\ref{sec:cat}
we discuss more refined semantics that are more suitable for richer
categories.

@@

The operational semantics define a relation rather than a function, but the
goal is for them to still be reversible; in particular, we wish for the
relation to mode check so that for any combinator {{c}} we can treat either
its left or its right side as an input; this allows us to define {{sym}} as
just an argument swap. We can then give an operational interpretation to the
semantic relation by defining two functions: a forward evaluator and a
backwards evaluator that exercise the relation in opposite directions. It is
straightforward to check that if the forward evaluation of {{c v1}} produces
{{v2}} as a possible answer then the backwards evaluation of {{c v2}}
produces {{v1}} as a possible answer.

In the interest of brevity, we have treated the type annotations in the base
isomorphisms as implicit; for example, the rule for {{assocr*}} should really
read:

@@

@@

discuss two generalizations of the semantics

vector spaces (more refined because it doesn't collpase 1/b to b; also each
vector space must have a zero vector; so we have an explicit notion of
``error'' which we can use and track at run time. Adding this error
eliminates at RUN TIME relations which are not isomorphisms I think.

a potentionally better and more semantics is based on groupoids; mention that
we can have sets (with stucture) who cardinality is fractionals and that this
would formalize nicely the negative information
